Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-11 doc/invt.tex
add backup figure scripts

\documentclass[reprint]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{hyperref}

@@

We now give an interesting deduction
following from the error functional
given in Eq. \eqref{eq:error_asym1}:
the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_k$'s,
but with different eigenvalues, $\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of numbers, $c_k \ge 0$,
%
\begin{align}
&
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t') \bigr)
\right)
%
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t') \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k q(t) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
This inequality follows from the fact
that the following quadratic polynomial of $Y$,
$$
\int_0^t
  dt' \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
%\ge 0,
$$
is always nonnegative, and hence
has a non-positive discriminant,
$B^2 - 4 \, A \, C$.
%
This condition furnishes the desired inequality.
%
Note that the last expression of \eqref{eq:CSineq}
depends on $q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t') \right) = c_k$
for all $k = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
%
$$
\alpha(t') = \frac{              1              }
                  { \lambda_k \, (t' + t_{k0} ) },
\quad
\mathrm{with\;\;}
t_{k0} = \frac{             t            }
              { e^{ \lambda_k q(t) } - 1 }.
$$
%
Such a solution is possible only if
%
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
  u_k(t') = c_k \left(t' + t_{k0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_k(t) = 1$ and $u_k(0) = e^{-\lambda_k \, q(t)}$,
  we have
  $$
  c_k = \frac{ 1 }{ t + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { t + t_{k0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_k \, q(t) + \lambda_k \, q(t')
  = \log c_k + \log\left( t' + t_{k0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_k \, \alpha(t') = \frac{ 1 } { t' + t_{k0} }.
  $$
}

Now an updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
Thus, in terms of asymptotic convergence,
the single-bin scheme
(as adopted in the WL algorithm)
is generally the most efficient
updating scheme.

We must, however, emphasize that the above result
holds only under the long time limit,
and it is not necessarily true
for finite length simulations.
%
For example, consider a system that is initially
prepared under a fixed $\alpha_0$
before starting the schedule of, $\alpha(t)$.
%
Then for shorter times, $t \approx 0$,
the error is given by Eq. \eqref{eq:error_eql},
which is clearly smaller if some eigenvalues
are less than $1.0$ (the single-bin case).
%
In practice, real simulations
may have long achieved the desired precision
before reaching the asymptotic regime,
especially with some $\lambda_k$'s close to $0$.

@@

Direct optimization of Eq. \eqref{eq:error_asym1}
for an arbitrary updating matrix $\mathbf w$
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a reasonably optimal $\alpha(t)$ with a simpler expression.
%
From Eq. \eqref{eq:alpha_opt},
we consider the following invert-time formula
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(t) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ t + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.

@@

Under the approximation, we have
$$
\frac{ \lambda_k^2 }{ \lambda \, (2 \, \lambda_k - \lambda) } \ge 1,
$$
with the equality achieved only at $\lambda = \lambda_k$,
we see the from Eq. \eqref{eq:error_asym_invt}
that the minimal error is achieved with
$\lambda_2 = \dots = \lambda_n = \lambda$,
agreeing with the general conclusion
in Sec. \ref{sec:optWL}.

In practice, however,
real simulations
may be relative short
and have not entered
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t) = 1/[\lambda (t + t_0)]$.

@@

We shall show that the optimal $\lambda$
depends on the simulation length,
and it tends to decrease as simulation lengthens.
%
Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{    \Gamma_k    }
       { 2 \, (t + t_0) }
  G\left( \nu_k, \frac{ t_0 } { t + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\nu_k
\,
\left(
  1
  +
  \frac{ 1 - r^{2 \, \nu_k - 1} }
       {   2 \, \nu_k - 1       }
\right).
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (t + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$, and the other around $\nu_k = 1$,
with the former being deeper albeit narrower.
%
This means that for a relatively short simulation,
hence a relatively large value of $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable as simulation lengthens,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$ under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error is always achieved at
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

For an arbitrary spectrum of eigenvalues,
$\lambda_2, \dots, \lambda_n$,
we expect that the optimal $\lambda$ to shift gradually to
smaller values as the simulation length, $t$, increasing.
%
Asymptotically, we expect the optimal $\lambda$
to be around $\min_k \lambda_k$.

@@

In practice, we usually use a homogeneous updating scheme
that is translational invariant.
%
We shall study these schemes below.

@@

Consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij}
  =
  \mu_{ |i - j| }
  + \mu_{ i + j - 1 }
  + \mu_{ 2 n + 1 - i - j },
  \label{eq:w_band}
\end{equation}
%
where $\mu_0, \dots, \mu_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
\mu_0 + 2 \, \mu_1 + \cdots + 2 \, \mu_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
The matrix is translational invariant under the reflective
boundary condition\cite{bussi2006}.
%
On the right-hand side of Eq. \eqref{eq:w_band},
the first term represents the relative updating magnitude
to a neighbor that is $|i - j|$ bins away,
whereas the next two terms are added
%according to the boundary condition
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006}.
%
Similarly, if $i$ represents a periodic variable,
the last two terms can be dropped, and
$|i-j|$ is evaluated
under the minimum image convention\cite{dama2014}.
This case is, however, not considered further.

@@

We can define a discrete kernel function from the set of numbers as
$$
\hat \mu_i = \mu_{ |i| }.
$$
Then, according to Eq. \eqref{eq:vt_diffeq_mbin},
the rate of change of the bias potential, $\dot v_i(t)$,
is given by a convolution of the above kernel, $\hat \mu_j$,
and the normalized histogram, $h_j/p_j$\cite{bussi2006},
and Eq. \eqref{eq:wband_eigenvalue1}
shows that the eigenvalues, $\lambda_k$, of the updating scheme
is the Fourier transform of the kernel.
%
So, for the updating scheme to be stable,
the Fourier transform of the kernel
must be nonnegative.

@@

Particularly, if $\mu_2 = \dots = \mu_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%%
%\begin{equation}
%\arraycolsep=3.6pt\def\arraystretch{1.4}
%\mathbf w
%=
%\left(
%  \begin{array}{cccccccc}
%    1 - \mu_1   & \mu_1 & 0 & \dots & 0 \\
%    \mu_1 & 1 - 2 \, \mu_1  & \mu_1 & \dots & 0 \\
%    \vdots & &  & & \vdots \\
%    0 & \dots & \mu_1 & 1 - 2 \, \mu_1  & \mu_1 \\
%    0 & \dots & 0 & \mu_1 & 1 - \mu_1
%  \end{array}
%\right).
%\label{eq:wnn}
%\end{equation}
%%
%We have
and the eigenvalues are
\begin{equation}
  \lambda_k = 1 - 4 \, \mu_1 \sin^2 \frac{(k - 1) \, \pi}{2n}.
\label{eq:wnn_eigenvalue}
\end{equation}
%
Thus, for this updating matrix to be stable,
we need
$\min \lambda_k = \lambda_n > 0$,
and
$$
\mu_1 < \frac{1}{4 \, \cos^2\frac{\pi}{2n} }.
$$

@@

For perfect sampling,
we have, from Eq. \eqref{eq:Gamma_perfect},
$\Gamma(p) = 1$, and
$$
\begin{aligned}
\Err_A
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, \mu_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, \mu_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
%
It is somewhat more convenient to change variable $p \to \frac{ \pi } { 2 } - p$,
and
$$
\begin{aligned}
\Err_A
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, \mu_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, \mu_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - \mu_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, \mu_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, \mu_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, \mu_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, \mu_1 - \lambda}}
{\sqrt{8 \mu_1}}.
$$
Thus,
$$
\begin{aligned}
\Err_A
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - \mu_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 \mu_1 - \lambda - 4 \mu_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \frac{ \left(1 - \mu_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 \mu_1 - \lambda - 4 \mu_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, \mu_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 \mu_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, \mu_1} { 1 - 2 \, \mu_1 },
\label{eq:lambda_nn_perfect}
\end{equation}
%
and the asymptotic error in the optimal case is given by
%
\begin{equation}
\Err_A
=
\frac{n}{t}
\left(
  1+ \frac{2 \, \mu_1^2}{1-4 \, \mu_1}
\right).
\label{eq:error_nn_prefect}
\end{equation}
%
%From Eq. \eqref{eq:error_nn}, it is clear that
Clearly, the error
increases with the magnitude of $\mu_1$
no matter its sign,
and the minimal value, obtained at $\mu_1 = 0$,
corresponds to the single-bin scheme.
%
Note that we have ignored the residual error
and assumed $\lambda < 2 \, \min \lambda_k = 2 - 8 \, \mu_1$
for all $k$ in the above solution.
%
Fortunately,
the $\lambda$ from Eq. \eqref{eq:lambda_nn_perfect}
satisfies this condition
by the stability condition,
$4 \, \mu_1 < 1$.

@@

>>>>

\documentclass[reprint, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{hyperref}

@@

The error functional given in Eq. \eqref{eq:error_asym1}
leads to an interesting deduction:
the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_k$'s,
but with different eigenvalues, $\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of numbers, $c_k \ge 0$,
%
\begin{align}
&
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t') \bigr)
\right)
%
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^t dt'
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t') \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k q(t) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
This inequality follows from the fact
that the following quadratic polynomial of $Y$,
$$
\int_0^t
  dt' \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
%\ge 0,
$$
is always nonnegative, and hence
has a non-positive discriminant,
$B^2 - 4 \, A \, C$.
%
Note that the last expression of \eqref{eq:CSineq}
depends on $q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t') \right) = c_k$
for all $k = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
%
$$
\alpha(t') = \frac{              1              }
                  { \lambda_k \, (t' + t_{k0} ) },
\quad
\mathrm{with\;\;}
t_{k0} = \frac{             t            }
              { e^{ \lambda_k q(t) } - 1 }.
$$
%
Such a solution is possible only if
%
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
  u_k(t') = c_k \left(t' + t_{k0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_k(t) = 1$ and $u_k(0) = e^{-\lambda_k \, q(t)}$,
  we have
  $$
  c_k = \frac{ 1 }{ t + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { t + t_{k0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_k \, q(t) + \lambda_k \, q(t')
  = \log c_k + \log\left( t' + t_{k0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_k \, \alpha(t') = \frac{ 1 } { t' + t_{k0} }.
  $$
}

Now an updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
Thus, in terms of asymptotic convergence,
the single-bin scheme
(as adopted in the WL algorithm)
is most efficient among updating schemes.

We must, however, emphasize that the above result
holds only under the long time limit,
and it is not necessarily true
for finite length simulations.
%
For example, consider a system that is initially
prepared under a fixed $\alpha_0$
before starting the schedule of, $\alpha(t)$.
%
Then for shorter times, $t \approx 0$,
the error is given by Eq. \eqref{eq:error_eql},
which is clearly smaller if some eigenvalues
are less than $1.0$ (the single-bin case).
%
In practice, real simulations
may have long achieved the desired precision
before reaching the asymptotic regime,
especially if the updating matrix
has some $\lambda_k$'s close to $0$.

@@

Direct optimization of Eq. \eqref{eq:error_asym1}
for an arbitrary updating matrix $\mathbf w$
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a reasonably optimal $\alpha(t)$ with a simpler expression.
%
From Eq. \eqref{eq:alpha_opt},
we consider the following invert-time formula
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

@@

Under the approximation, we have
$$
\frac{ \lambda_k^2 }{ \lambda \, (2 \, \lambda_k - \lambda) } \ge 1,
$$
with the equality achieved only at $\lambda = \lambda_k$,
and the minimal error from \eqref{eq:error_asym_invt}
is achieved with
$\lambda_2 = \dots = \lambda_n = \lambda$,
agreeing with the general conclusion
in Sec. \ref{sec:optWL}.

Real simulations, however,
can be relatively short
and need not enter
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(t) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ t + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium, or saturated, error of mode $k$
at the terminal value
of $\alpha(t) = 1/[\lambda (t + t_0)]$.

@@

We now show that the optimal $\lambda$
tends to decrease with the simulation length.
%
Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{    \Gamma_k    }
       { 2 \, (t + t_0) }
  G\left( \nu_k, \frac{ t_0 } { t + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\nu_k
\,
\left(
  1
  +
  \frac{ 1 - r^{2 \, \nu_k - 1} }
       {   2 \, \nu_k - 1       }
\right).
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (t + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$,
and the other shallower one around $\nu_k = 1$.
%
This means that for a relatively short simulation,
hence a relatively large value of $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$, or increasing simulation length,
under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error is always achieved at
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

For an arbitrary spectrum of eigenvalues,
$\lambda_2, \dots, \lambda_n$,
we expect that the optimal $\lambda$ to shift gradually to
smaller values as the simulation length, $t$, increasing.
%
Asymptotically, we expect the optimal $\lambda$
to be around $\min \lambda_k$.

@@

We now consider a class of commonly-used updating scheme
for sampling a uniform distribution, $p_i = 1/n$.
%
These schemes are homogeneous,
or translational invariant with respect to $z$.
%
This invariance is natural for a periodic variable\cite{dama2014}.
%
Below, however, we focus on a variable
under the reflective boundary conditions.

@@

Consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij}
  =
  \mu_{ |i - j| }
  + \mu_{ i + j - 1 }
  + \mu_{ 2 n + 1 - i - j },
  \label{eq:w_band}
\end{equation}
%
where $\mu_0, \dots, \mu_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
\mu_0 + 2 \, \mu_1 + \cdots + 2 \, \mu_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
The matrix is translational invariant under the reflective
boundary condition\cite{bussi2006}.
%
On the right-hand side of Eq. \eqref{eq:w_band},
the first term represents the relative updating magnitude
to a neighbor that is $|i - j|$ bins away,
whereas the next two terms are added
%according to the boundary condition
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006}.
%
\note{Similarly, if $i$ represents a periodic variable,
the last two terms can be dropped, and
$|i-j|$ is evaluated
under the minimum image convention\cite{dama2014}.
This case is, however, not considered further.}

@@

We can define a discrete kernel function from the set of numbers as
$$
\hat \mu_i = \mu_{ |i| }.
$$
Then, according to Eq. \eqref{eq:vt_diffeq_mbin},
the rate of change of the bias potential, $\dot v_i(t)$,
is given by the convolution of the above kernel, $\hat \mu_j$,
and the normalized histogram, $h_j/p_j$\cite{bussi2006},
and Eq. \eqref{eq:wband_eigenvalue1}
shows that the eigenvalues, $\lambda_k$, of the updating scheme
is the Fourier transform of the kernel.
%
So, for the updating scheme to be stable,
the Fourier transform of the kernel
must be nonnegative.

@@

Particularly, if $\mu_2 = \dots = \mu_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%%
%\begin{equation}
%\arraycolsep=3.6pt\def\arraystretch{1.4}
%\mathbf w
%=
%\left(
%  \begin{array}{cccccccc}
%    1 - \mu_1   & \mu_1 & 0 & \dots & 0 \\
%    \mu_1 & 1 - 2 \, \mu_1  & \mu_1 & \dots & 0 \\
%    \vdots & &  & & \vdots \\
%    0 & \dots & \mu_1 & 1 - 2 \, \mu_1  & \mu_1 \\
%    0 & \dots & 0 & \mu_1 & 1 - \mu_1
%  \end{array}
%\right).
%\label{eq:wnn}
%\end{equation}
%%
%We have
and the eigenvalues are
\begin{equation}
  \lambda_k = 1 - 4 \, \mu_1 \sin^2 \frac{(k - 1) \, \pi}{2n}.
\label{eq:wnn_eigenvalue}
\end{equation}
%
Thus, for this updating matrix to be stable,
we need
$\min \lambda_k = \lambda_n > 0$,
or
$$
\mu_1 < \frac{1}{4 \, \cos^2\frac{\pi}{2n} }.
$$

@@

For perfect sampling,
we have, from Eq. \eqref{eq:Gamma_perfect},
$\Gamma(p) = 1$, and
$$
\begin{aligned}
\Err_A
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, \mu_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, \mu_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
%
It is somewhat more convenient to change variable $p \to \frac{ \pi } { 2 } - p$,
and
$$
\begin{aligned}
\Err_A
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, \mu_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, \mu_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - \mu_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, \mu_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, \mu_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, \mu_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, \mu_1 - \lambda}}
{\sqrt{8 \mu_1}}.
$$
Thus,
$$
\begin{aligned}
\Err_A
&=
\frac{      n       }
     { \lambda \, t }
\left(
 \frac{ 2 - 4 \, \mu_1 - \lambda }
      {          4               }
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - \mu_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 \mu_1 - \lambda - 4 \mu_1 z^2) }
\right)
\\
&=
\frac{       n      }
     { \lambda \, t }
\left(
  \frac{ 2 - 4 \, \mu_1 - \lambda }
       {          4               }
 +
 \frac{ \left(1 - \mu_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
      { 2 - 4 \mu_1 - \lambda - 4 \mu_1 z_{\pm}^2                 }
\right)
\\
&=
\frac{       n      }
     { \lambda \, t }
\left(
  \frac{ 2 - 4 \, \mu_1 - \lambda }
       {          4               }
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 \mu_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, \mu_1} { 1 - 2 \, \mu_1 },
\label{eq:lambda_nn_perfect}
\end{equation}
%
and the asymptotic error in the optimal case is given by
%
\begin{equation}
\Err_A
=
\frac{n}{t}
\left(
  1+ \frac{2 \, \mu_1^2}{1-4 \, \mu_1}
\right).
\label{eq:error_nn_prefect}
\end{equation}
%
%From Eq. \eqref{eq:error_nn}, it is clear that
Clearly, the error
increases with the magnitude of $\mu_1$
no matter its sign,
and the minimal value, obtained at $\mu_1 = 0$,
corresponds to the single-bin scheme.
%
Note that we have ignored the residual error
and assumed $\lambda < 2 \, \min \lambda_k = 2 - 8 \, \mu_1$
for all $k$ in the above solution.
%
Fortunately,
the $\lambda$ from Eq. \eqref{eq:lambda_nn_perfect}
satisfies this condition
by the stability condition,
$4 \, \mu_1 < 1$.
\note{This is because,
$$
\frac{ 1 - 4 \, \mu_1 } { 1 - 2 \, \mu_1 } < 2 - 8 \, \mu_1
\quad \leftrightarrow \quad
1 < 2 - 4 \, \mu_1.
$$
}

@@

\section{\label{eq:stabilize_wband}
Stabilization}
